---
title: "Workflow urban form and urban metrics"
output: html_document
---

For detailed explanations and comments on what happens in each step, please refer to the respective code file and the additional files in the respective code and process data folders.

Functionality to be included in the future:
- parcel, block, building IDs indicating relations between each other
- open source implementation of code currently run using ArcGIS Pro / arcpy (mainly building metrics toolbox)


# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r imports, warning=FALSE}
library(jsonlite)
library(qgisprocess)
library(xml2)
library(reticulate) # for arcpy
library(glue)
library(fs)
library(stringr)
library(lubridate)
```

```{r paths, results='hide'}

# to run for test data
test <- FALSE

# create environment that contains all variables defined here so they can be accessed when running child scripts
child_env <- new.env()

# define directory paths
child_env$scriptdir <- getwd()
child_env$maindir <- dirname(dirname(getwd()))
child_env$basedata <- normalizePath("../../A_basedata", winslash = "/")
child_env$procdata <- normalizePath("../B2_procdata", winslash = "/")
child_env$resultdata <- normalizePath("../../A_resultdata", winslash = "/")

# set dir paths to use for test data
if (test){
  child_env$procdata <- paste0(child_env$procdata, "_test")
  child_env$resultdata <- paste0(child_env$resultdata, "_test")
  child_env$test <- TRUE
}

# set parameters
child_env$commit_sha <- system("git rev-parse HEAD", intern = TRUE)  # get SHA of latest commit to save code versions next to outputs
child_env$date <- format(Sys.Date(), "%Y%m%d")
child_env$CRS <- 32632  # project crs: integer from valid EPSG code

# include defined variables in global environment as well
list2env(as.list(child_env), envir = .GlobalEnv)  

# set chunk output dir (needed for knit_child)
knitdir <- dir_create(file.path(scriptdir, 'B0_R_knit'))  # -> added to gitignore
knitr::opts_knit$set(output.dir = knitdir)  

```


## Arcpy setup

(for loading data from gdbs and running the building metrics toolbox)

```{r arcpy-setup}
# Point reticulate to the correct ArcGIS Pro Python
use_python("C:/Program Files/ArcGIS/Pro/bin/Python/envs/arcgispro-py3/python.exe", required = TRUE)

# Attach arcpy inside child_env
child_env$arcpy <- import("arcpy", delay_load = TRUE)
```
>> When getting an error, go into Tools > Global Options > Python > Uncheck box "Automatically activate ..." 
>> R will restart > run imports and path definitions again

>> For RuntimeError: The Product License has not been initialized., open ArcGIS Pro and log in.


## QGIS Process Setup

- copies all QGIS model files into the QGIS system processing folder so they can be called as algorithms 
- also copies files containing 'add_to_toolbox' in their names to the toolbox as custom scripts so they can be called from models (currently only 00.02.04aa)
- sometimes old versions of algorithms get loaded here although they run correctly in QGIS, try:
    - run qgis_configure(use_cached_data = TRUE) and delete the cache folder, then rerun this code
    
```{r qgisprocess-setup}
model_path <- file.path(Sys.getenv("APPDATA"), "QGIS", "QGIS3", "profiles", "default", "processing", "models")
script_path <- file.path(Sys.getenv("APPDATA"), "QGIS", "QGIS3", "profiles", "default", "processing", "scripts")

if (!file.exists(model_path) | !file.exists(script_path)) {
  
  print("Processing models or script folder not in default location, in QGIS Settings > Options > Processing, check for the correct folder paths.")
  
} else {
  
  # get all QGIS models and scripts from project scripts folder
  model_files <- list.files(path = scriptdir, pattern = "\\.model3$", recursive = TRUE, full.names = TRUE)
  script_files <- list.files(path = scriptdir, pattern = "add_to_toolbox", recursive = TRUE, full.names = TRUE)
  
  # Edit each file to have the file name set as the algorithm name -> sometimes have old names, easier access when same as file name
  for (model_file in model_files) {
    model_name <- tools::file_path_sans_ext(basename(model_file))

    # Read and modify XML
    xml <- read_xml(model_file)
    option_nodes <- xml_find_all(xml, './/Option[@name="model_name"]')

    if (length(option_nodes) > 0) {
      xml_set_attr(option_nodes[[1]], "value", model_name)
      write_xml(xml, model_file)
    }
  }
  
  file.copy(from = model_files, to = file.path(model_path, basename(model_files)), overwrite = TRUE)
  cat(length(model_files), "model3 files copied to", model_path, "\n")
  
  file.copy(from = script_files, to = file.path(script_path, basename(script_files)), overwrite = TRUE)
  cat(length(script_files), "custom script file copied to", model_path, "\n")
  
  # Load and view added models
  qgis_configure(quiet = TRUE)
  qgis_search_algorithms(provider = "script|model")
}
```

## Writing and running Children

- Don't use chunk names in children, they will result in problems when running a script several time from the parent and when they are not unique between all children
- Include `knitr::opts_knit$set(root.dir = scriptdir)` in the child script so the relative paths work -> this only works when actually knitting (i.e. running from parent), not when working in the child itself. When working in the child and running individual chunks, you can use `setwd(scriptdir)` in the chunks where you are defining paths. You can either use `normalizePath()` to make the paths absolute when defining them or keep them relative and also run `setwd()` in the chunk where you are calling the paths. 
- We are using a child environment for variable interaction between parent and child. Variables can be defined in the parent using `child_env$varname` and then be called as `varname` in the child. Variables defined within the child can again be accessed from the parent using `child_env$childvar`. To avoid large variables staying in memory longer than necessary, use `rm(list = c("childvar1", "childvar2"))` at the end of your children.
  - Children need to be called specifying `child_env`: 
    - Use `knitr::knit_child(spath("step1.Rmd"), envir = child_env, quiet = TRUE)` to run `.rmd` files (quiet suppresses the progress bar and messages)
    - Use `sys.source("step1.R", envir=child_env)` to run `.R` files
- Assign the command calling the child to a variable (`knittedoutput`) to not have the full child code and results printed as a long string
    - You can view the knitted output of a child using `cat(knittedoutput)` (better format than `print()`)
- If you want to get specific output from the child, you can assign it to variables (e.g. `message`) and then print them using `child_env$message` in the parent. 
    - You can get the message from a command using e.g. `message <- capture.output(st_write(overlap, outpath, append = FALSE))` 


Helper functions for calling scripts:

```{r script-model-access}
# uses subdir defined in each Step, if the subdir has more subfolders, include them in the filename using / 
spath <- function(fname){
  normalizePath(file.path(scriptdir, child_env$subdir, fname))
}

# run QGIS algorithms:
run_model <- function(algorithm, rawjson) {

  # convert JSON copied from QGIS to list
  inputparams <- jsonlite::fromJSON(rawjson)

  # run the QGIS model
  qgis_result <- qgis_run_algorithm(
    algorithm,
    !!!inputparams
  )

  return(qgis_result)
}

# write a txt file containing the sha key and date into the result and processdata dir to reference the code version the file was created from
write_sha <- function() {
  dirs <- file.path(c(resultdata, procdata), child_env$subdir)
  
  # create dirs if they don't exist
  lapply(dirs, function(d) if (!dir.exists(d)) dir.create(d, recursive = TRUE))
  
  # write file with commit version to dir
  paths <- file.path(dirs, "commitversion.txt")
  invisible(lapply(paths, function(p) writeLines(c(date, commit_sha), normalizePath(p, mustWork = FALSE))))
}
```

# Code

## --- 01 Area definition ---
(Community selection necessary for feature selection in further harmonisation steps)

```{r sub01}
child_env$subdir <- "01_AreaDefinition"
write_sha()
```

### 01a Get isochrones

- Gets isochrone in 45 minute range.

```{r}
# due to changes in the road network, this does not always reproduce the same isochrones -> change the name of the version you want to keep manually and specify it in the next chunk to keep using it

child_env$ors_url <- "http://url/ors"
sys.source(spath("01a_Isochrones.R"), envir=child_env)
```


### 01b Community Selector based on Threshold value

- Selects municipalities within the isochrone to be used as study area. 

```{r, warning=FALSE, echo=FALSE}
child_env$isochrone_version <- "_20240606"  # set to isochrone version to use

knittedoutput <- knitr::knit_child(spath("01b_Community_Selector_Threshold.Rmd"), envir = child_env, quiet = TRUE)
print(child_env$message)
```


## --- 00 Harmonisation ---

- Harmonises German and French data sets into a common version. 

### 00a Municipalities

* Municipality harmonisation included in script 01b Community selection.


### 00b 01 Addresses

```{r sub00b-01}
child_env$subdir <- "00_Harmonisation/00b_01_addresses"
write_sha()
```

```{r}
sys.source(spath("00b_01_address_csv_to_gpkg.R"), envir=child_env)
```

### 00b 02 Buildings

```{r sub00b-02}
child_env$subdir <- "00_Harmonisation/00b_02_buildings"
child_env$lodyear <- "2020"  # the german buildings are stored in gdbs for different years on our server -> set the year of german buildings to load into the project folder 
write_sha()
```

#### 00b.02.01 Copy buildings for germany to gpkg from server
```{r}
sys.source(spath("00b_02_01_loadfromserver_DE.R"), envir=child_env)
```

#### 00b.02.02 Prepare data for join
```{r}
sys.source(spath("00b_02_02a_prepareforjoin_crs_extent.R"), envir=child_env)
```

##### DE
```{r}
# change field names if necessary 
sys.source(spath("00b_02_02ba_prepareforjoin_fieldcalculation_DE.R"), envir=child_env)
```

##### FR
```{r}
algorithm <- "model:00b_02_02bb_prepareforjoin_fieldcalculation_FR"

#for (dep in c('57', '67')) {  # for testing
for (dep in c('57', '67', '54', '68', '88')) {
  rawjson <- glue('{{
    "departements": "{procdata}/00_Harmonisation/00b_02_buildings/00b_02_02_prepareforjoin/departements.gpkg",
    "new_fields": "{procdata}/00_Harmonisation/00b_02_buildings/00b_02_02_prepareforjoin/00b_02_02b_fieldcalculation/fr{dep}_parts.gpkg",
    "reprojected_and_clipped_building_parts_fr": "{procdata}/00_Harmonisation/00b_02_buildings/00b_02_02_prepareforjoin/00b_02_02a_clip_communitiesbuffer20/fr{dep}_parts.gpkg"
  }}')
    
  cat("\n---\nProcessing: FR", dep, format(now(), "%H:%M:%S"), "\n")
  print(rawjson)
  
  qgis_result <- run_model(algorithm, rawjson)
  print(qgis_result)
}
```

#### 00b.02.03 Join FR & DE
```{r}
sys.source(spath("00b_02_03_join_frde.R"), envir=child_env)
```

#### 00b.02.04a Parts to buildings by region
```{r}
algorithm <- "model:00b_02_04a_parts_to_buildings_byregion"
buildingsdir <- file.path(procdata, child_env$subdir, "00b_02_04_parts_to_building", "00b_02_04a_buildings")
fieldcalcdir <- file.path(procdata, child_env$subdir, "00b_02_02_prepareforjoin", "00b_02_02b_fieldcalculation")
dir_create(buildingsdir)

regions <- dir_ls(fieldcalcdir) %>%  path_file() %>%  str_extract("^[^_]+") #extracts everything before the first underscore from files in field calc folder to get region names list

for (region in regions) {
  buildingspath <- file.path(procdata, "00_Harmonisation/00b_02_buildings/00b_02_04_parts_to_building/00b_02_04a_buildings", paste0(region, "_buildings.gpkg"))
  print(buildingspath)
  rawjson <- glue('{{
    "buildings": "{buildingspath}",
    "national_data_sources": "{procdata}/00_Harmonisation/00b_02_buildings/00b_02_03_join_frde/00b_02_03_parts.gpkg",
    "osm": "{procdata}/00_Harmonisation/00b_02_buildings/00b_02_04a_OSM_merged_BW_RP_HE_FR_20km.gpkg",
    "region": "{region}"  
  }}')
  
  cat("\n---\nProcessing:", region, format(now(), "%H:%M:%S"), "\n")
  print(rawjson)
  
  qgis_result <- run_model(algorithm, rawjson)
  print(qgis_result)
}
```

#### 00b.02.04b Merge regions back together
```{r}
sys.source(spath("00b_02_04b_merge_regions_back_together.R"), envir=child_env)
```

#### 00b.02.05 Join addresses to buildings
```{r}
sys.source(spath("00b_02_05_join_addr_to_bldg.R"), envir=child_env)
```

#### 00b.02.06 Clip to communities
```{r}
sys.source(spath("00b_02_06_clip_to_communities.R"), envir=child_env)
```


### 00c Parcels

```{r sub00c-01}
child_env$subdir <- "00_Harmonisation/00c_parcels"
write_sha()
```

```{r}
sys.source(spath("00c_parcel_harmonisation.R"), envir=child_env)
```


### 00d Streets

```{r sub00d-01}
child_env$subdir <- "00_Harmonisation/00d_streets"
write_sha()
```

```{r}
sys.source(spath("00d_street_harmonisation.R"), envir=child_env)
```


### 00e Corine

```{r sub00d-01}
child_env$subdir <- "00_Harmonisation/00e_corine"
write_sha()
```

```{r}
algorithm <- "model:00e_CorineHarmonisation"

area_selection <- file.path(resultdata, '01_AreaDefinition', '01b_communities.gpkg')
corine <- paste0(file.path(basedata, 'EU', 'Corine', 'clc2018', 'corine_2018.gdb'),"|layername=clc18_all")
corine_selection <- file.path(resultdata, '00_Harmonisation', '00e_corine', '00e_corine.gpkg')

rawjson <- glue('{{
    "area_selection": "{area_selection}",
    "corine": "corine",
    "corine_selection": "{corine_selection}"
  }}')

qgis_result <- run_model(algorithm, rawjson)
```


## --- 02c Development Block Identification --- 

```{r sub02c}
child_env$subdir <- "02c_DevelopmentBlockIdentification"
write_sha()
```

### 02c.01 Parcel Analysis

- Prepare parcel data.

```{r}
sys.source(spath("02c_01_ParcelAnalysis.Rmd"), envir=child_env)
```

### 02c.02 Delaunay Triangulation, Block Dissolve

- Run block delimitation separately for Germany and France

```{r}
algorithm <- "model:02c_02_DelaunayTriangulation_BlockDissolve"

# FR
rawjson <- '{
    "devblocks_by_country": "{resultdata}/02c_DevelopmentBlockIdentification/02c_02_DevelopmentBlockIdentificationFR.gpkg",
    "join_parcel_street_names": "{resultdata}/02c_DevelopmentBlockIdentification/02c_01_DevelopmentBlockIdentificationFR.gpkg"
  }'

qgis_result1 <- run_model(algorithm, rawjson)

# DE
rawjson <- '{
    "devblocks_by_country": "{resultdata}/02c_DevelopmentBlockIdentification/02c_02_DevelopmentBlockIdentificationDE.gpkg",
    "join_parcel_street_names": "{resultdata}/02c_DevelopmentBlockIdentification/02c_01_DevelopmentBlockIdentificationDE.gpkg"
  }'

qgis_result2 <- run_model(algorithm, rawjson)
```

### 02c.03 Merge and Assign Block ID

- Merge and Assign Block ID

```{r}
sys.source(spath("02c_03_Merge_BlockID.Rmd"), envir=child_env)
```


## --- 03 Building Type Identification ---

- Random forest building type classification based on calculation of building metrics.

```{r sub03}
child_env$subdir <- "03_BuildingTypeIdentification"
write_sha()
```

### 03.1 Building Metrics

#### 03.1a Data to GeoDatabase

- Takes harmonised data sets needed for building metric calculation and saves them into a GDB for better handling in ArcGIS Pro.

```{r}  
sys.source(spath("03_01a_x5_data_to_gdb.R"), envir=child_env)
```

#### 03.1b Run Toolbox

- Building metrics calculation using ArcGIS Pro toolbox.
- Takes a long time.

```{r}
sys.source(spath("03_01b_run_toolbox.R"), envir=child_env)
```

### 03.2 Prepare Training Data

- Prepares training data from different sources and harmonises it to our building geometries and common building types.

```{r}
knittedoutput <- knitr::knit_child(spath("03_02_clean_training_data.Rmd"), envir = child_env, quiet = TRUE)
print(child_env$message1)
print(child_env$message2)
```


### 03.3 Random Forest

- Trains and runs a random forest model to classify buildings.

```{r}
knittedoutput <- knitr::knit_child(spath("03_03_randomforest.Rmd"), envir = child_env, quiet = TRUE)
print(child_env$message)
print(child_env$rf_model)
child_env$importance_table
child_env$importance_plot
```


### 03.4 Type to blocks

- Join Building type with largest volume to development blocks

```{r}
knittedoutput <- knitr::knit_child(spath("03_04_type_to_blocks.Rmd"), envir = child_env, quiet = TRUE)
```

## --- 04 Block Age GHSAge ---

- Extracts block construction age from GHSAge layer.

```{r sub04}
child_env$subdir <- "04_BlockAge_GHSAGE"
write_sha()
```


### 04 Extract GHS Age 

- Extract block age from GHS Age layer 
- Tests different extraction approaches, by default chooses approach with highest accuracy within tolerance to apply to whole dataset; define `approach` parameter to override

```{r}
child_env$approach <- "wmean"

knittedoutput <- knitr::knit_child(spath("04_extract_ghs_age.Rmd"), envir = child_env, quiet = TRUE)
#child_env$results
cat(child_env$message1)
cat(child_env$message2)
```


## --- 05a Accessibility Analysis --- 

- Calculate Closeness Centrality of each block based on distance-weighted reachable population

This part of the code is based on and largely taken from [Jehling & Kluwe (2026)](https://doi.org/10.5281/zenodo.19481958)
It mainly uses python and QGIS scripts.

```{r sub05a}
child_env$subdir <- "05a_AccessibilityAnalysis"
write_sha()
```

### 05a.01 Centrality 50 Prozent

- Loads street network from OSM. Projection of a point grid of any grid width onto a road network, adding the number of inhabitants to points.

> run 05a_01_OSM_highways2points+einwohner.py from QGIS Python console

### 05a.02 OSM Highways2Points + Einwohner

- Selection of 50 % of all points (less computation time), requests to local ORS instance - origin-distance matrix is created from each starting point to all target points.

> Run 05a_02_centrality_50Prozent.py from QGIS Python console

### 05a.03 Accessibility Calculations

#### 05a.03a Calculate closeness centrality

- Merge all matrices, apply decay functions and weighting for each point, eliminate extreme values. Calculate composite index from driving times and walking/cycling distances.
- merge_matrix, new_accessibility, interpolation (bools) are set to TRUE -> change in script if running again and no new calculations are necessary

```{r}
sys.source(spath("05a_03a_impedance+weighting_50perc.R"), envir=child_env)
```


#### 05a.03b Join regions 

- Join to blocks and join regions together

```{r}
# join to blocks and join regions together
sys.source(spath("05a_03b_join_regions.R"), envir=child_env)
```

